---
title: "Cyclistic Bike-Sharing Data Explorations"
author: "Paw Hermansen"
date: "2022-09-03"
output:
  pdf_document: default
  html_document: default
subtitle: Google Data Analytics Capstone (case study)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r install libraries, include = FALSE}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
if(!require(fpeek)) install.packages("fpeek", repos = "http://cran.us.r-project.org")
```

```{r use libraries, include = FALSE}
library(tidyverse)
library(lubridate)
library(fpeek)

options(dplyr.summarise.inform=F)
```


# Scenario and the Business Task

Cyclistic is a bike-share company in Chicago. Cyclistic has two kinds of customers, casual riders who pay per ride or per day, and members who pay annually.

Cyclistic wants to convert more casual riders into annual members and as part of this they need to know how annual members and casual riders use Cyclistic bikes differently.

The Business Task is:

> Analyze the available dataset to answer how annual members and casual riders use Cyclistic bikes differently.

# Getting the Data

The data is downloaded from <https://divvy-tripdata.s3.amazonaws.com/index.html> and has the form of 12 zipped csv files---one file for each month of the period August 2021 to July 2022, both included.

The data is public and made available by Motivate International Inc. under this [license](https://www.divvybikes.com/data-license-agreement). Any information about riders has been removed. This means that this data can be used without having concerns about data privacy but it also means that we cannot include for example the riders sex, age, place of residence, number of members, and number of unique casual riders in the analysis which surely would have been a big source of very useful information.

The data seems trustworthy. As far as I can assess the data is original, data is available for a continues period over several years including very current data and I believe it contains all the rides. Also, it is linked-to by the Google Data Analytics Certification Capstone assignment.

Because the data contains the full set of rides in the examined period for all riders and all geographies, or in short the full population, it has no problems with selection bias---except from the possibility that missing values are skew, for example if casual riders have missing many more values than members. I will look at missing values and the possibility of skewness in missing values later in this report.

```{r unzip, results='hide'}
# Unzip the zipped files
all_zipfilenames <- list.files("data", pattern="\\d{6}-divvy-tripdata\\.zip$", ignore.case=TRUE, full.names=TRUE)
sapply(all_zipfilenames, unzip, exdir="data", overwrite=TRUE)
```

Counting the number of rows in the csv files show that it's close to six million rows in total. Because of this I decide to do the analysis in R.

```{r find all csv filenames}
# List the names of the csv files
all_csvfilenames <- list.files("data", pattern="\\d{6}-divvy-tripdata\\.csv$", ignore.case=TRUE, full.names=TRUE)
```

```{r find number of rows}
# Find the number of rows in each csv file
number_of_csvfile_rows <- list()
total_num_rows <- 0
for (filename in all_csvfilenames) {
  num_rows <- peek_count_lines(filename)
  total_num_rows <- total_num_rows + num_rows
  cat(sprintf("%s, number of rows: %6d\n", filename, num_rows))
  number_of_csvfile_rows <- append(number_of_csvfile_rows, num_rows)
}
cat("Total number of rows = ", total_num_rows, "\n")
```

Visually inspecting the top lines of two of the files show that they both have a header as the first row and have the data start in the second row.

```{r show top rows of a csv file}
peek_head("data/202108-divvy-tripdata.csv", n = 3)
```

```{r show top rows of another csv file}
peek_head("data/202207-divvy-tripdata.csv", n = 3)
```

The following check confirms that all the files have a header row and that the header is the same for all the files. This means that I expect the csv files to have columns of the same type of data in the same order. It is also noted that some of the data values are missing (when there is nothing between the commas).

```{r all csv files have same header}
header_rows <- sapply(all_csvfilenames, readLines, n=1)
cat("Number of different header rows =", length(unique(header_rows)))
```

Finally the csv files are read and combined into a data frame named `tripdata_df`, or more exactly into a R tibble but I'll just call it a data frame in the rest of this report.

```{r reading}
tripdata_df <- do.call(rbind, lapply(all_csvfilenames, read_csv, col_types=cols()))

glimpse(tripdata_df)
```

In all we have `r nrow(tripdata_df)` rows for the twelve months.

# Data Integrity and Data Cleaning

## Check of Data Loading

To be sure we have loaded the data correctly I find the number of rides in each month of the loaded data and compare them with the number of rows in the csv files found earlier. It is seen that every csv file has exactly one row extra which is the header row present in all the csv files. This means that all rows of the csv files have been loaded.

```{r check data loaded ok}
tripdata_df %>%
  mutate(date_yearmon = format(started_at, "%Y-%m")) %>%
  group_by(date_yearmon) %>%
  summarise(number_of_rows_in_df = n()) %>%
  add_column(number_of_rows_in_csvfiles = unlist(number_of_csvfile_rows)) %>%
  mutate(difference = number_of_rows_in_csvfiles - number_of_rows_in_df)
```

## ride_id

A unique key of R type Character (text string) for each ride. All rows have a different `ride_id` and they are all 16 characters long.

```{r ride_id}
cat("Number of missing values: ", sum(is.na(tripdata_df$ride_id) | tripdata_df$ride_id == ""), "\n")
```

```{r count distinct values}
cat("Number of duplets: ", nrow(tripdata_df) - n_distinct(tripdata_df$ride_id), "\n")
```

```{r length(s) of ride_id}
tripdata_df %>% group_by(str_length(ride_id)) %>% summarize(count = n())
```

## rideable_type

Three different types of bikes are present: `classic_bike`, `docked_bike`, `electric_bike`. No row is missing this value (or it would have shown-up as a fourth value in the following code-chunk and the three counts would not have summed up to the total number of rows).

Because this is categorial data, i.e. only a few fixed different values exist, I change its type in `tripdata_df` to the R type factor.

```{r the different values}
table(tripdata_df$rideable_type)
```

```{r change type of rideable_type}
tripdata_df$rideable_type <- as.factor(tripdata_df$rideable_type)
str(tripdata_df$rideable_type)
```

## started_at, ended_at

Start-time and end-time of each ride of type date-time. All rows have both `started_at` and `ended_at`.

```{r check start-time and end-time for missing values}
cat("Number of missing started_at: ", sum(is.na(tripdata_df$started_at)), "\n")
cat("Number of missing ended_at:   ", sum(is.na(tripdata_df$ended_at)), "\n")
```

## start_station_name, end_station_name, start_station_id, end_station_id

Because about 15% of the station names and station ids are missing I decide to not use the station data and I delete the columns.

```{r stations}
num_rows = nrow(tripdata_df)
missing_ssn <- sum(is.na(tripdata_df$start_station_name) | tripdata_df$start_station_name == "")
missing_esn <- sum(is.na(tripdata_df$end_station_name) | tripdata_df$end_station_name == "")
missing_ssi <- sum(is.na(tripdata_df$start_station_id) | tripdata_df$start_station_id == "")
missing_esi <- sum(is.na(tripdata_df$end_station_id) | tripdata_df$end_station_id == "")

cat(sprintf("Number of missing start_station_names: %6d = %.1f%%\n", missing_ssn, 100 * missing_ssn / num_rows))
cat(sprintf("Number of missing end_station_names:   %6d = %.1f%%\n", missing_esn, 100 * missing_esn / num_rows))
cat(sprintf("Number of missing start_station_ids:   %6d = %.1f%%\n", missing_ssi, 100 * missing_ssi / num_rows))
cat(sprintf("Number of missing end_station_ids:     %6d = %.1f%%\n", missing_esi, 100 * missing_esi / num_rows))
```

```{r delete station columns}
tripdata_df <- select(tripdata_df, -c(start_station_name, end_station_name, start_station_id, end_station_id))
glimpse(tripdata_df)
```

## start_lat, start_lng

The geographical latitude and longitude of type R `double`, i.e. a number, of the trip start point. No values are missing.

```{r start geographical point}
cat("Number of missing start_lat: ", sum(is.na(tripdata_df$start_lat)), "\n")
cat("Number of missing start_lng: ", sum(is.na(tripdata_df$start_lng)), "\n")
```

## end_lat, end_lng

The geographical latitude and longitude of type R `double`, i.e. a number, of the trip end point. `r sum(is.na(tripdata_df$end_lat) | is.nan(tripdata_df$end_lng))` values are missing.

```{r end geographical point}
cat("Number of missing end_lat: ", sum(is.na(tripdata_df$end_lat)), "\n")
cat("Number of missing end_lng: ", sum(is.na(tripdata_df$end_lng)), "\n")
```

It turns out that the two values are missing in exactly the same rows so `r sum(is.na(tripdata_df$end_lat) | is.nan(tripdata_df$end_lng))` rows in all are affected. That is about `r sprintf("%.1f", 100*sum(is.na(tripdata_df$end_lat) | is.nan(tripdata_df$end_lng)) / nrow(tripdata_df))`% of all the rows. This number is so small that deleting the rows with missing values will not change the results of this analyses in any visible way. So I delete the involved `r sum(is.na(tripdata_df$end_lat) | is.nan(tripdata_df$end_lng))` rows.

If the missing values are in rows that for example are exactly of one kind of `member_casual` then deleting these rows will introduce skewness and bias but as shown the maximal possible bias is too small to influence the results in this analyses.

```{r missing end geographical points}
cat("Number of at least one missing: ", sum(is.na(tripdata_df$end_lat) | is.na(tripdata_df$end_lng)), "\n")
```

```{r delete rows with missing end coordinates}
tripdata_df <- tripdata_df %>% filter(!is.na(tripdata_df$end_lat) & !is.na(tripdata_df$end_lng))
```

## member_casual

Two different possibilities are present: `casual`, `member`. No row is missing this value (or it would have shown-up as a third value in the following code-chunk and the two counts would not have summed up to the total number of rows).

Because this is categorical data, i.e. only a few fixed different values exist, I change its type in `tripdata_df` to the R type factor.

```{r the different member_casual values}
levels(tripdata_df$member_casual)
```

```{r change type of member_casual}
tripdata_df$member_casual <- as.factor(tripdata_df$member_casual)
str(tripdata_df$member_casual)
```

# Explorations Into the Differences Between Casual Riders and Members

## Number of Rides by Casual Riders and Members

```{r number of rides}
cat(sprintf("Number of Rides by Casual Riders: %d  %.0f%%\n", nrow(tripdata_df[tripdata_df$member_casual == 'casual', ]), 100 * nrow(tripdata_df[tripdata_df$member_casual == 'casual', ]) / nrow(tripdata_df)))
cat(sprintf("Number of Rides by Members      : %d  %.0f%%\n", nrow(tripdata_df[tripdata_df$member_casual == 'member', ]), 100 * nrow(tripdata_df[tripdata_df$member_casual == 'member', ]) / nrow(tripdata_df)))
```

The data available doesn't give any information about the number of member or different casual riders and so I cannot make any conclusions about how often the two different rider types go for a ride.

## Seasonal

During summer months approximately half of the riders are casual rides and half are members. Far fewer rides are done during the Chicago winter months and especially the casual riders are not riding during winter months. During January and February more than 80% of the rides are done by members.

```{r seasonal}
tripdata_df %>%
  ggplot(aes(x=as.Date(started_at), fill=member_casual)) +
  geom_histogram(bins=50, position = "dodge") +
  scale_y_continuous(labels=scales::comma_format(big.mark=',', decimal.mark = '.')) +
  scale_x_date(date_breaks="1 month", date_labels="%b") +
  theme(axis.text.x=element_text(hjust = 0)) +
  labs(title="Number of Rides Over a Year", subtitle="Split between Casual Riders and Members", x="", y="Number of Rides", fill="")
```

```{r seasonal winter}
tripdata_df %>%
  filter(month(started_at) %in% c(1,2)) %>%
  group_by(member_casual) %>% summarize(number_of_rides_in_Jan_and_Feb=n()) %>% mutate(number_of_rides_percents=round(100 * number_of_rides_in_Jan_and_Feb / sum(number_of_rides_in_Jan_and_Feb)))
```

```{r seasonal summer}
tripdata_df %>%
  filter(month(started_at) %in% c(5,6,7,8)) %>%
  group_by(member_casual) %>% summarize(number_of_rides_in_May_to_Aug=n()) %>% mutate(number_of_rides_percents=round(100 * number_of_rides_in_May_to_Aug / sum(number_of_rides_in_May_to_Aug)))
```

## Days of the Week

During weekends more than half the rides are done by casual riders while the opposite is true during weekdays. This trend mostly happens because casual riders ride more during weekends than during the rest of the week while members ride more evenly over the full week, though with a very small tendency to ride less during weekends.

Because casual riders are so much more inactive during the winter months, this trend must be strongest during the summer months.

```{r weekly}
tripdata_df %>%
  group_by(weekday=wday(started_at, label=TRUE, week_start=1), member_casual) %>%
  summarize(number_of_rides=n()) %>%
  mutate(number_of_rides_prop = number_of_rides / sum(number_of_rides)) %>%
  ggplot(aes(x=weekday, y=number_of_rides, fill=member_casual)) + geom_bar(stat="identity") +
  geom_text(position=position_stack(vjust = 0.5), aes(label=scales::percent(number_of_rides_prop, accuracy=1)), alpha=0.6) +
  scale_y_continuous(labels=scales::comma_format(big.mark=',', decimal.mark = '.')) +
  labs(title="Number of Rides Each Weekday", subtitle="Split between Casual Riders and Members", x="", y="Number of Rides", fill="")
```

## Rideable Type

Members mostly ride a classical bike while casual riders are split evenly between clissical bikes and electrical bikes with a small group on docked bikes.

```{r bike types}
tripdata_df %>%
  group_by(member_casual, rideable_type) %>%
  summarize(number_of_rides=n()) %>%
  mutate(number_of_rides_prop = number_of_rides / sum(number_of_rides)) %>%
  ggplot(aes(x=member_casual, y=number_of_rides, fill=rideable_type)) + geom_bar(stat="identity") +
  geom_text(position=position_stack(vjust = 0.5), aes(label=scales::percent(number_of_rides_prop, accuracy=1)), alpha=0.6) +
  scale_y_continuous(labels=scales::comma_format(big.mark=',', decimal.mark = '.')) +
  labs(title="Number of Casual Riders and Members", subtitle="Split between Different Bike Types", x="", y="Number of Rides", fill="")
```

## Time of Day

Members have a very clear pattern showing many rides in the hours just before and after work-hours on weekdays, Monday to Friday. Casual riders show a noticeable smaller peak right after work-hours and only a nearly invisible peak right before work-hours

During weekends neither casual riders nor members show such peaks.

```{r time of day}
tripdata_df %>%
  mutate(part_of_week = case_when(wday(started_at) %in% c(1,7) ~ "Weekend", TRUE ~ "Weekday")) %>%
  mutate(time_of_day = 3600*hour(started_at) + 60*minute(started_at) + second(started_at)) %>%
  ggplot(aes(x=time_of_day)) + geom_histogram(bins=96) +
  facet_grid(part_of_week ~ member_casual) + aes(fill=member_casual) +
  scale_x_continuous(breaks = c(0, 6*3600, 12*3600, 18*3600, 24*3600), labels = c("12am", "6am", "12pm", "6pm", "12am")) +
  scale_y_continuous(labels=scales::comma_format(big.mark=',', decimal.mark = '.')) +
  labs(title="Number of Rides During the Day", subtitle="Split between Casual Riders and Members, and between Weekend or not",
       x="Time Of Day", y="Number of Rides", fill="")
```

## Duration of Rides

```{r duration calc}
durationdata_df <- tripdata_df %>%
  mutate(ride_duration_minutes = as.double(difftime(ended_at, started_at, units="mins")))
```

```{r duration min and max }
cat("Ride duration in minutes:\n")
cat(sprintf("Min    = %6.1f minutes\n", min(durationdata_df$ride_duration_minutes)))
cat(sprintf("Max    = %6.1f minutes\n", max(durationdata_df$ride_duration_minutes)))
```

It is seen that some ride durations are negative which of course shows that some of the duration data is wrong.

```{r count less than and zero}
cat("Number of Duration < 0:", durationdata_df %>% filter(ride_duration_minutes < 0) %>% nrow(), "\n")
cat("Number of Duration = 0:", durationdata_df %>% filter(ride_duration_minutes == 0) %>% nrow(), "\n")
```

Because this is such a small number compared to the total number of rows, I remove all rows with duration less or equal to zero.

```{r remove rows with negative durations}
durationdata_df <- durationdata_df %>% filter(0 < ride_duration_minutes)

cat("Number of Rows After Removing Rows with Non-Positive Durations:", nrow(durationdata_df))
```

```{r duration stats}
cat("Ride duration in minutes:\n")
cat(sprintf("Min    = %6.1f minutes\n", min(durationdata_df$ride_duration_minutes)))
cat(sprintf("Max    = %6.1f minutes\n", max(durationdata_df$ride_duration_minutes)))
cat(sprintf("Mean   = %6.1f minutes\n", mean(durationdata_df$ride_duration_minutes)))
cat(sprintf("Median = %6.1f minutes\n", median(durationdata_df$ride_duration_minutes)))
cat(sprintf("99%% percentile = %.1f minutes\n", quantile(durationdata_df$ride_duration_minutes, probs=0.99)))
```

The max ride duration time is very much larger than the mean which again is larger than the median and this usually implies that the duration times are skewed with most short times and fewer larger and larger times. This is confirmed in the following chart where I cut-off the durations to 110 minutes because 99% of all durations are less than 10.8 minutes.

```{r duration chart}
durationdata_df %>% 
  filter(as.double(ride_duration_minutes) < 110) %>%
  arrange(ride_duration_minutes) %>%
  ggplot(aes(x=ride_duration_minutes)) + geom_histogram(bins=200) +
  facet_grid(. ~ member_casual) + aes(fill=member_casual) +
  scale_y_continuous(labels=scales::comma_format(big.mark=',', decimal.mark = '.')) +
  labs(title="Number of Rides of Different Durations", subtitle="Split between Casual Riders and Members",
       x="Ride Duration in Minutes", y="Number of Rides", fill="")

```

From the chart and from the following table of percentiles it is seen that members overall ride shorter rides than casual rides. The median ride duration is only about 9 minutes for members whereas it is about 14 minutes for casual riders.

```{r duration stats split on member and casual}
durationdata_df %>%
  group_by(member_casual) %>%
  summarise(Min = min(ride_duration_minutes),
            Percentile20 = quantile(ride_duration_minutes, probs = .2),
            Percentile50 = quantile(ride_duration_minutes, probs = .5),
            Mean = mean(ride_duration_minutes),
            Percentile80 = quantile(ride_duration_minutes, probs = .8),
            Percentile90 = quantile(ride_duration_minutes, probs = .9),
            Percentile99 = quantile(ride_duration_minutes, probs = .99),
            Max = max(ride_duration_minutes))
```

# Presentation

## Slide 1: Business Task / Purpose

**How Does Members and Casual Riders use Cyclistic Bikes Differently?**

- author: Paw Hermansen
- date: Sept 11, 2022

## Slide 2: Data

- The data is downloaded from <https://divvy-tripdata.s3.amazonaws.com/index.html> for the period August 2021 to July 2022, both included.
- The data is public and made available by Motivate International Inc. under a public license.
- All information about riders has been removed, i.e. no information about for example the frequency of rides for the two user types.
- The Data needed some clean-up but was otherwise good.

## Slide 3: Seasonal Differences

- During summer months approximately half of the riders are casual rides and half are members. 
- During winter months more than 80% of the rides are done by members.

```{r presentation chart year, echo=FALSE}
tripdata_df %>%
  ggplot(aes(x=as.Date(started_at), fill=member_casual)) +
  geom_histogram(bins=50, position = "dodge") +
  scale_y_continuous(labels=scales::comma_format(big.mark=',', decimal.mark = '.')) +
  scale_x_date(date_breaks="1 month", date_labels="%b") +
  theme(axis.text.x=element_text(hjust = 0)) +
  labs(title="Number of Rides Over a Year", subtitle="Split between Casual Riders and Members", x="", y="Number of Rides", fill="")
```

## Slide 4: Weekly Differences

- During weekends more than half the rides are done by casual riders
- During weekdays more than half the rides are done by members


```{r presentation chart week, echo=FALSE}
tripdata_df %>%
  group_by(weekday=wday(started_at, label=TRUE, week_start=1), member_casual) %>%
  summarize(number_of_rides=n()) %>%
  mutate(number_of_rides_prop = number_of_rides / sum(number_of_rides)) %>%
  ggplot(aes(x=weekday, y=number_of_rides, fill=member_casual)) + geom_bar(stat="identity") +
  geom_text(position=position_stack(vjust = 0.5), aes(label=scales::percent(number_of_rides_prop, accuracy=1)), alpha=0.6) +
  scale_y_continuous(labels=scales::comma_format(big.mark=',', decimal.mark = '.')) +
  labs(title="Number of Rides Each Weekday", subtitle="Split between Casual Riders and Members", x="", y="Number of Rides", fill="")
```

## Slide 4: Daily Differences

- Members very clearly rides frequently in the hours just before and after work-hours on weekdays, Monday to Friday.
- Casual riders show this a lot less especially nearly no peak shows before work-hours.


```{r presentation chart day, echo=FALSE}
tripdata_df %>%
  mutate(part_of_week = case_when(wday(started_at) %in% c(1,7) ~ "Weekend", TRUE ~ "Weekday")) %>%
  mutate(time_of_day = 3600*hour(started_at) + 60*minute(started_at) + second(started_at)) %>%
  ggplot(aes(x=time_of_day)) + geom_histogram(bins=96) +
  facet_grid(part_of_week ~ member_casual) + aes(fill=member_casual) +
  scale_x_continuous(breaks = c(0, 6*3600, 12*3600, 18*3600, 24*3600), labels = c("12am", "6am", "12pm", "6pm", "12am")) +
  scale_y_continuous(labels=scales::comma_format(big.mark=',', decimal.mark = '.')) +
  labs(title="Number of Rides During the Day", subtitle="Split between Casual Riders and Members, and between Weekend or not",
       x="Time Of Day", y="Number of Rides", fill="")
```


## Slide 5: Ride Durations

- For members the 20% shortest rides takes less than about 4 and a half minutes
- For casual riders the 20% shortest rides takes less than about 7 minutes


| Duration in Minutes for Percentiles | 20%  | 50%  | 80%  |
|-------------------------------------|------|------|------|
| Casual Rider                        |  7.1 | 14.9 | 30.6 |
| Member                              |  4.6 |  9.0 | 18.0 |


## Slide 6: Summary

- Casual riders ride far less than members during winter months
- Casual riders ride more during weekends and members ride more during weekdays
- Casual riders rides are longer than member's rides.

Also worth mentioning:

- Casual riders do not ride to and from work as much as members appear to do. The difference is largest in the morning (before work-hours).


